---
title: "Milestone 4"
author: "Group 6"
date: "2022-09-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(ggthemes)
require(extrafont)
require(matlib)
require(reshape)
require(corrplot)
require(tidyverse)
require(mapdata)
require(car)
```

```{r}
riverquality<-read.csv("../Data/ms4cleaned.csv")
riverquality<-riverquality[,-1]
rq08<-read.csv("../Data/rq08.csv")
rq90<-read.csv("../Data/rq90.csv")
rq98<-read.csv("../Data/rq98.csv")

num.time<-riverquality[,c("period","Phosphorus","Nitrogen","E.Coli","Turbidity")]
num.time$period <- as.factor(num.time$period)

num.melt <- melt(data=num.time,
id.vars = "period",
variable.name = "Compound")


class.new <- num.time[1500,1]
x.new <- num.time[1500,-1]

num.time <- num.time[-45, ]
num.melt <- melt(data=num.time,
id.vars = "period",
variable.name = "Compound")
```

# Mahalanobis distance 

The distance between x1500 and the compound model is 2.250291, and the dataset has a probability of 0.69 falling into this distance (Fig. 1). 

```{r, fig.cap= "Distribution of the Mahalanobis distance of river observations to the aggregated model"}


est.mu <- colMeans(num.time[, -1])
est.covar <- var(num.time[, -1])

(d.new <- mahalanobis(x.new, center = est.mu, cov = est.covar))
pchisq(d.new, df = 4, lower.tail = FALSE)


dM <- mahalanobis(num.time[,-1], center = est.mu, cov = est.covar)
upper.quantiles <- qchisq(c(.9, .95, .99), df=4)
density.at.quantiles <- dchisq(x=upper.quantiles, df=4)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)
ggplot(data.frame(dM), aes(x=dM)) +
geom_histogram(aes(y=..density..), bins=nclass.FD(dM),
fill="white", col="black") +
stat_function(fun="dchisq", args = list(df=4),
col="blue", size=2, alpha=.7, xlim=c(0,65)) +
geom_point(aes(x=d.new, y=0), size=3, col="red") +
geom_segment(data=cut.points,
aes(x=upper.quantiles, xend=upper.quantiles,
y=rep(0,3), yend=density.at.quantiles),
col="red", size=1) +xlab("Mahalanobis distances and cut points") +
ylab("Histogram and density")


```

The new measure is consistent with both the first and second time periods; 1990-2017 and 1998-2017, and closest to 1998-2017.

```{r}
est.mu1 <- colMeans(subset(num.time, period=="1990-2017")[,-1])
est.covar1 <- var(subset(num.time, period=="1990-2017")[,-1])
est.mu2 <- colMeans(subset(num.time, period=="1998-2017")[,-1])
est.covar2 <- var(subset(num.time, period=="1998-2017")[,-1])
est.mu3 <- colMeans(subset(num.time, period=="2008-2017"
)[,-1])
est.covar3 <- var(subset(num.time, period=="2008-2017"
)[,-1])

(d.new1 <- mahalanobis(x.new, center = est.mu1, cov = est.covar1))
pchisq(d.new1, df = 4, lower.tail = FALSE)
(d.new2 <- mahalanobis(x.new, center = est.mu2, cov = est.covar2))
pchisq(d.new2, df = 4, lower.tail = FALSE)
(d.new3 <- mahalanobis(x.new, center = est.mu3, cov = est.covar3))
pchisq(d.new3, df = 4, lower.tail = FALSE)
```

# Cleaned data (overall)
We can see that all river health indicators; E.Coli, Phosphorus, Nitrogen and Turbidity peak at a value of around 110 (Fig. 2). Phosphorus, Nitrogen and turbidity also show smaller peaks at 250 and 340 (Fig. 2). 
```{r equality, echo=FALSE}
waterq<-read.csv("../Data/ms4cleaned.csv", header=T)
wq<-data.frame(waterq$num_id, waterq$E.Coli,waterq$Phosphorus, waterq$Nitrogen,waterq$Longitude, waterq$lat, waterq$Turbidity)
summary(wq)
```

```{r plotting, fig.cap="Histograms for river health indicators; E.Coli, Nitrogen, Phosphorus and Turbidity in NZ rivers", echo=FALSE}
par(mfrow=c(2,2))
hist(waterq$E.Coli)
hist(waterq$Nitrogen)
hist(waterq$Phosphorus) 
hist(waterq$Turbidity)

```

```{r, fig.cap= "Boxplots showing the distributions of NZ river health and information variables", echo=FALSE}
waterq.melt<-melt(waterq)
waterq$num_id<-as.factor(waterq$num_id)
ggplot(data=waterq.melt, aes(factor(variable), value)) + 
geom_boxplot() + 
facet_wrap(~variable, scale="free")
```

In this case, we fit the river health indicators; "E.Coli" with "Turbidity" to assign a plot of Fitted vs Residual values. 

```{r Fit, fig.cap="Plot of residuals vs. fitted values and normal Q-Q plot for E.Coli and Turbidity factors in NZ rivers", echo=FALSE}
mod1 <- lm(waterq$E.Coli ~ as.factor(waterq$Turbidity), data = waterq)

mod1resid <- mod1$res
mod1fitted <- mod1$fit
par(mfrow=c(1,2))
plot(mod1resid~mod1fitted, xlab="Fitted values", ylab="Residuals") 
qqnorm(mod1resid,main="")
qqline(mod1resid) 
```

Residual vs fitted values are not normally distributed about zero, with differing 'heights' (Fig. 4). This suggests non-constant variance. The Normal Q-Q plot shows that the data is not normally distributed, as it does not show a straight line (Fig. 4).

Levene Test: $H_0:$ equal variance $H_1:$ not all variance are equal

```{r Levene Test}
leveneTest(waterq$E.Coli ~ as.factor(waterq$Turbidity), data = waterq)
```

Levene's test produces the F-statistic = 9.0905, and p-value = 2.2e-16, which provides strong evidence to reject the null hypothesis of equal variance.

We have the fitted data that has no constant variance and is not distributed normally. Therefore, we use permutation test instead of ANOVA.

Permutation: $H_0:$ There is no difference in E.Coli across Turbidity $H_1:$ E.Coli differs across Turbidity

```{r Permutation}
Fobs<-anova(mod1)$F[1]
Fnull<-rep(NA,2000)
for(t in 1:2000)
{
reorder<-sample(waterq$Turbidity)  
Fnull[t]<-(anova(lm(waterq$E.Coli~as.factor(reorder))))$F[1]
}
p<-sum(Fnull>=Fobs)/2000
p
```

p-value = 0, therefore there is evidence that E.Coli differs across Turbidity.

# Period 1 (1990-2017)
E.Coli, Nitrogen and Turbidity show peaks at values around 110, whereas Phosphorus peaks at a high value of 350 (Fig. 5). This concentration of the E.Coli and Phosphorus data towards these value is evident in the boxplots (Fig. 6). We can also see a range of measurements of Nitrogen and Turbidity in NZ rivers during this period (1990-2017) (Fig. 5, Fig, 6). 
```{r equality 90, echo=FALSE}
waterq90<-read.csv("../Data/rq90.csv", header=T)
wq90<-data.frame(waterq90$num_id, waterq90$E.Coli,waterq90$Phosphorus, waterq90$Nitrogen,waterq90$Longitude, waterq90$lat, waterq90$Turbidity)
summary(wq90)
```

```{r plotting 90, fig.cap="Histograms showing NZ river health indicators; E.Coli, Nitrogen, Phosphorus and Turbidity from 1990-2017", echo=FALSE}
par(mfrow=c(2,2))
hist(waterq90$E.Coli)
hist(waterq90$Nitrogen)
hist(waterq90$Phosphorus)
hist(waterq90$Turbidity)
```

```{r, fig.cap="Boxplots showing the distributions of NZ river health and information variables from 1990-2017", echo=FALSE}
waterq90.melt<-melt(waterq90)
waterq90$num_id<-as.factor(waterq90$num_id)
ggplot(data=waterq90.melt, aes(factor(variable), value)) + 
geom_boxplot() + 
facet_wrap(~variable, scale="free")
```

Fit E.Coli with Turbidity to analyse the difference in all 3 periods.

```{r Fit 90, fig.cap="Plot of residuals vs. fitted values and normal Q-Q plot for E.Coli and Turbidity factors in NZ rivers; 1990-2017", echo=FALSE}
mod2 <- lm(waterq90$E.Coli ~ as.factor(waterq90$Turbidity), data = waterq90)

mod2resid <- mod2$res
mod2fitted <- mod2$fit
par(mfrow=c(1,2))
plot(mod2resid~mod2fitted, xlab="Fitted values", ylab="Residuals") 
qqnorm(mod2resid,main="")
qqline(mod2resid) 
```

Residual vs fitted values are not normally distributed about zero, with differing 'heights'. This suggests non-constant variance. The Normal Q-Q plot shows that the data is not normally distributed, as it does not show a straight line.

Levene Test: $H_0:$ equal variance $H_1:$ not all variance are equal

```{r Levene Test 90}
leveneTest(waterq90$E.Coli ~ as.factor(waterq90$Turbidity), data = waterq90)
```

The F-statistic = 1.7753, and p-value = 0.008519 of  Levene's test provides strong evidence to reject the null hypothesis of equal variance.

We have the fitted data that has no constant variance and is not distributed normally. Therefore, we use permutation test instead of ANOVA to test the equality of means.

Permutation: $H_0:$ There is no difference in E.Coli across Turbidity $H_1:$ E.Coli differs across Turbidity 

```{r Permutation 90, echo=FALSE}
Fobs<-anova(mod2)$F[1]
Fnull<-rep(NA,2000)
for(t in 1:2000)
{
reorder<-sample(waterq90$Turbidity)  
Fnull[t]<-(anova(lm(waterq90$E.Coli~as.factor(reorder))))$F[1]
}
p<-sum(Fnull>=Fobs)/2000
p
```

p-value = 0, therefore, there is evidence that E.Coli differs across turbidity in NZ rivers, from 1990-2017. 
# Period 2 (1998-2017)
We can see that there tends to be similar trends in river health indicators to those from 1990-2017. However, E.Coli shows a greater range of measurements and frequencies of all indicators are higher (Fig. 8, Fig. 9). 
```{r equality 98, echo=FALSE}
waterq98<-read.csv("../Data/rq98.csv", header=T)
wq98<-data.frame(waterq98$num_id, waterq98$E.Coli,waterq98$Phosphorus, waterq98$Nitrogen,waterq98$Longitude, waterq98$lat, waterq98$Turbidity)
summary(wq98)
```

```{r plotting 98, fig.cap="Histograms for river health indicators; E.Coli, Nitrogen, Phosphorus and Turbidity in NZ rivers from 1998-2017", echo=FALSE}
par(mfrow=c(2,2))
hist(waterq98$E.Coli)
hist(waterq98$Nitrogen)
hist(waterq98$Phosphorus)
hist(waterq98$Turbidity)
```

```{r, fig.cap="Boxplots of the distributions of NZ river health and information variables from 1998-2017", echo=FALSE}
waterq98.melt<-melt(waterq98)
waterq98$num_id<-as.factor(waterq98$num_id)
ggplot(data=waterq98.melt, aes(factor(variable), value)) +
geom_boxplot() + 
facet_wrap(~variable, scale="free")
```

Fit E.Coli with Turbidity to analyse the difference in all 3 periods.

```{r Fit 98, fig.cap="Plot of residuals vs. fitted values and normal Q-Q plot for E.Coli and Turbidity factors in NZ rivers; 1998-2017", echo=FALSE}
mod3 <- lm(waterq98$E.Coli ~ as.factor(waterq98$Turbidity), data = waterq98)

mod3resid <- mod3$res
mod3fitted <- mod3$fit
par(mfrow=c(1,2))
plot(mod3resid~mod3fitted, xlab="Fitted values", ylab="Residuals") 
qqnorm(mod3resid,main="")
qqline(mod3resid) # qqplot shows the fitted data has no normal distribution 
```

Residual vs fitted values are not normally distributed about zero, with differing 'heights'. This suggests non-constant variance (Fig. 10). The Normal Q-Q plot shows that the data is not normally distributed, as it does not show a straight line (Fig. 10). 

Levene Test: $H_0:$ equal variance $H_1:$ not all variance are equal

```{r Levene Test 98}
leveneTest(waterq98$E.Coli ~ as.factor(waterq98$Turbidity), data = waterq98)
```

The F-statistic = 5.1587, and p-value = 2.2e-16, Levene's test shows strong evidence to reject the null hypothesis of equal variance.

We have the Fitted data that has no constant variance and is not distributed normally. Therefore we use permutation test instead of anova.

Permutation: $H_0:$ There is no difference in E.Coli across Turbidity $H_1:$ E.Coli differs across Turbidity

```{r Permutation 98}
Fobs<-anova(mod3)$F[1]
Fnull<-rep(NA,2000)
for(t in 1:2000)
{
reorder<-sample(waterq98$Turbidity)  
Fnull[t]<-(anova(lm(waterq98$E.Coli~as.factor(reorder))))$F[1]
}
p<-sum(Fnull>=Fobs)/2000
p
```

p-value = 0, therefore there is evidence that E.Coli differs across turbidity in NZ rivers, throughout the period of 1998-2017. 

# Period 3 (2008-2017)
This period tends to show lower frequencies of all river health indicators (Fig. 11, Fig. 12)
```{r equality 08, echo=FALSE}
waterq08<-read.csv("../Data/rq08.csv", header=T)
wq08<-data.frame(waterq08$num_id, waterq08$E.Coli,waterq08$Phosphorus, waterq08$Nitrogen,waterq08$Longitude, waterq08$lat, waterq08$Turbidity)
summary(wq08)
```

```{r plotting 08, fig.cap="Histograms for river health indicators; E.Coli, Nitrogen, Phosphorus and Turbidity in NZ rivers from 2008-2017", echo=FALSE}
par(mfrow=c(2,2))
hist(waterq08$E.Coli)
hist(waterq08$Nitrogen)
hist(waterq08$Phosphorus)
hist(waterq08$Turbidity)

```

```{r, fig.cap="Boxplots of the distributions of NZ river health and information variables from 2008-2017", echo=FALSE}
waterq08.melt<-melt(waterq08)
waterq08$num_id<-as.factor(waterq08$num_id)
ggplot(data=waterq08.melt, aes(factor(variable), value)) + 
geom_boxplot() + 
facet_wrap(~variable, scale="free")
```

Fit E.Coli with Turbidity to analysis the difference in all 3 periods.

```{r Fit 08, fig.cap="Plot of residuals vs. fitted values and normal Q-Q plot for E.Coli and Turbidity factors in NZ rivers; 2008-2017", echo=FALSE}
mod4 <- lm(waterq08$E.Coli ~ as.factor(waterq08$Turbidity), data = waterq08)

mod4resid <- mod4$res
mod4fitted <- mod4$fit
par(mfrow=c(1,2))
plot(mod4resid~mod4fitted, xlab="Fitted values", ylab="Residuals") 
qqnorm(mod4resid,main="")
qqline(mod4resid) # qqplot shows the fitted data has no normal distribution 
```

Residual vs fitted values are not normally distributed about zero, with differing 'heights'. This suggests non-constant variance (Fig. 13). The Normal Q-Q plot shows that the data is not normally distributed, as it does not show a straight line (Fig. 13). 

Levene Test: $H_0:$ equal variance $H_1:$ not all variance are equal

```{r Levene Test 08}
leveneTest(waterq08$E.Coli ~ as.factor(waterq08$Turbidity), data = waterq08)
```

The F-statistic = 4.0892, and p-value = 4.218e-14, Levene's test shows strong evidence to reject the null hypothesis of equal variance.

Fitted data shows no homogeneity of variances and is not distributed normally. Therefore, for equality testing, we use permutation test instead of ANOVA.

Permutation: $H_0:$ There is no difference in E.Coli across Turbidity $H_1:$ E.Coli differs across Turbidity

```{r Permutation 08}
Fobs<-anova(mod4)$F[1]
Fnull<-rep(NA,2000)
for(t in 1:2000)
{
reorder<-sample(waterq08$Turbidity)  
Fnull[t]<-(anova(lm(waterq08$E.Coli~as.factor(reorder))))$F[1]
}
p<-sum(Fnull>=Fobs)/2000
p
```

p-value = 0, therefore, there is evidence that E.Coli differs across turbidity, between 2008 and 2017. 

Thus, we found evidence that E.Coli differs across turbidity in NZ rivers, across all three periods of time; 1990-2017, 1998-2017, 2008-2017. 
